A lottery model of density-dependent selection in evolutionary genetics

### 2017-05-05 manuscript.tex
latest

@@

%Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.

@@

Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Population persistence is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.

\begin{figure}
\centering
\includegraphics[scale=1]{grimeschematic.pdf}
\caption{\label{fig:grimeschematic} Schematic of Grime's triangle. The two axes show increasing levels of environmental stress and disturbance, respectively. Population persistence is not possible if the combination of stress and disturbance is too large (dashed line). This creates a triangle, each corner of which corresponds to a ``primary strategy''.} 
\end{figure}

@@

Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; section ``Primary strategies and Grime's triangle''). This represents a ``sanity check'' on our density-dependent lottery model. The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}).

@@

Here we only consider the last of these factors. To account for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article. Thus, we consider how selection alone shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.

Under Eq. \eqref{eq:master}, mutations which change $b$ or $d$ by the same  proportion (i.e. $b_j=(1+\epsilon) b_i$ or $d_j=(1-\epsilon) d_i$ where $j$ is the mutant type) have identical fitness effects (Appendix C). When propagule density is high  $L\gg 1$, the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$. When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral. Thus, selection under Eq. \eqref{eq:master}, which applies in ideal and stressed environments, does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.

The situation in disturbed environments, where mortality also affects juveniles, is more complicated. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. Thus, proportional changes in the remaining genetic part of mortality can have a tiny impact on $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.

Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental extreme schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes. The ideal environment lies at the origin (no stress or disturbance), while the stressful and disturbed environments lie at the limits of survival on their respective axes. The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance. The direction of evolutionary change is different at each vertex, leading to the emergence of different trait clusters or ``primary strategies''.

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The high-density (HD) interpretation of stress is also shown. The vertices of the triangles correspond to different environmental extremes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}

@@

%two general mechanisms by which fluctuating selection promotes coexistence. The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment. The storage effect occurs when some individuals are protected from selection; in the lottery model a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration. Protection from selection promotes coexistence in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only a fraction of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more adults from the abundant type die than can possibly be replaced by the rare types) \citep{chesson_1981}.

%The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects). If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not. This promotes coexistence even in the absence of frequency-dependent selection \citep{yi_2013}.

@@

>>>>

\section*{Introduction}

@@

@@

Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:axes}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Population persistence is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} Schematic of Grime's triangle and the associated direction of selection in our model. The two axes show increasing levels of environmental stress and disturbance, respectively. Population persistence is not possible if the combination of stress and disturbance is too large (dashed line). This creates a triangle, each corner of which corresponds to an environmental extreme and associated ``primary strategy''. The high-density (HD) interpretation of stress is shown as a rotation of the triangle with respect to density. Vertical arrows show how selection acts on $b$, $c$, and $d$: $c$ is irrelevant at low $L$, but otherwise $b$, $c$ and $d$ all experience the same selective force.} 
\end{figure}

@@

Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environmental extremes (section ``Primary strategies and Grime's triangle''). This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model in different extreme cases.

@@

Accounting for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article. We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects. Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C). Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits .

Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$. However, beyond this, selection in our density-dependent lottery model does not favor one trait improvement over another (Fig. 3). This applies regardless of the initial magnitude of $b$ or $d$. More specifically, in ideal and stressed environments, where Eq. \eqref{eq:master} applies, $b$ and $d$ mutations of size $\epsilon$ have identical fitness effects (Appendix C), and the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$.

In disturbed environments, $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily swamped by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality. In Appendix C, we argue that $b$ and $d$ mutations should have comparable effects on fitness (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves and so $d$ cannot run ahead of $b$ for long (diminishing returns).

@@

@@

The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}).
